---
title: "nsink"
output: github_document
---
<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

<!-- badges: start -->
[![R build status](https://github.com/jhollist/nsink/workflows/R-CMD-check/badge.svg)](https://github.com/jhollist/nsink/actions)
[![Codecov test coverage](https://codecov.io/gh/jhollist/nsink/branch/main/graph/badge.svg)](https://codecov.io/gh/jhollist/nsink?branch=main)
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.6341565.svg)](https://doi.org/10.5281/zenodo.6341565)
[![DOI](https://joss.theoj.org/papers/10.21105/joss.04039/status.svg)](https://doi.org/10.21105/joss.04039)
<!-- badges: end -->

# Statement of need

The `nsink` package is an R implementation of the methods described in [Kellogg et. al (2010)](https://doi.org/10.1016/j.ecoleng.2010.02.006).  Previous implementation of this approach relied on a manual, vector based approach that was time consuming to prepare.  This approach uses a hybrid raster-vector approach that takes relatively little time to set up for each new watershed and relies on readily available data. Total run times vary, but range from minutes up to 5 hours depending on options selected.  Previous versions took weeks of manual data manipulation.  Thus, `nsink` was developed to satisfy the need for quicker implementation of the NSink method as described in [Kellogg et. al (2010)](https://doi.org/10.1016/j.ecoleng.2010.02.006).  

# `nsink` functionality
As of `r lubridate::today()` user functions for the `nsink` package are:

- `nsink_get_huc_id()`: A function for searching the name of a USGS Watershed Boundary Dataset Hydrologic Unit (<https://www.usgs.gov/core-science-systems/ngp/national-hydrography/watershed-boundary-dataset>) and retrieving its 12-digit Hydrologic Unit Code (HUC).  
- `nsink_get_data()`: Using any acceptable HUC ID (e.g. 2-digit to 12-digit), this function downloads the NHDPlus, SSURGO, NLCD Land Cover, and the NLCD Impervious for that HUC.   
- `nsink_prep_data()`: `nsink` needs data in a common coordinate reference system, from mutliple NHDPlus tables, and from different portions of SSURGO.  This function completes these data preparation steps and outputs all data, clipped to the HUC boundary.
- `nsink_calc_removal()`: Quantifying relative N removal across a landscape is a key aspects of an `nsink` analysis.  The `nsink_calc_removal()` function takes the object returned from `nsink_prep_data()` and calculates relative N removal for each landscape sink.  See Kellogg et al [-@kellogg2010geospatial] for details on relative N removal estimation for each sink.
- `nsink_generate_flowpath()`: This function uses a combination of flow determined by topography, via a flow-direction raster, for the land-based portions of a flow path and of downstream flow along the NHDPlus stream network.   
- `nsink_summarize_flowpath()`: Summarizing removal along a specified flow path requires relative N removal and a generated flow path.  This function uses these and returns a  summary of relative N removal along a flow path for each sink. 
- `nsink_generate_static_maps()`: This function analyzes N removal at the watershed scale by summarizing the results of multiple flow paths. Four static maps are returned: 1)removal efficiency; 2)loading index; 3)transport index; 4)delivery index.  Removal efficiency is a rasterized version of the `nsink_calc_removal()` output.  Loading index is N sources based on NLCD categories.   Transport index is a heat map with the cumulative relative N removal along flow paths originating from a grid of points, density set by the user, across a watershed, highlighting the gradient of downstream N retention. Delivery index is the result of multiplying the loading index and the transport index, and shows potential N delivery from different sources, taking into account the relative N removal as water moves downstream. 
- `nsink_plot()`: A function that plots each raster in the list returned from `nsink_generate_static_maps()`.   
- `nsink_build()`: One of the drivers behind the development of the `nsink` package was to provide `n-sink` analysis output that could be used more broadly (e.g. within a GIS).  The `nsink_build()` runs a complete `nsink` analysis and outputs R objects, shapefiles and/or TIFFs.
- `nsink_load()`: Essentially the inverse of the `nsink_build()` function, this function takes a folder of files, likely created by `nsink_build()`, and reads them into R.

# Installation instructions

At this time we plan on maintaining the `nsink` package as a GitHub only package and thus it won't be available directly from CRAN.  You may use the `install_github()` function from the `remotes` package to install it.  The code below will take care of installing `remotes` and installing `nsink` from the GitHub repository.

```{r, eval=FALSE}
install.packages("remotes")
remotes::install_github("usepa/nsink", dependencies = TRUE, build_vignettes = TRUE)
```

And then to load up the package:

```{r eval=FALSE}
library(nsink)
```

# Documentation and examples

All functions are documented, with examples, and that documentation may be accessed, in R, via the usual help functions.  Additionally, an introduction to the `nsink` package with a more detailed workflow is documented in a vignette.  

```{r eval=FALSE}
# Load up package
library(nsink)

# Access package level help
help(package = "nsink")

# Access the Introduction to nsink vignette
vignette("intro", package = "nsink")
```

# Contributing

If you would like to contribute to the `nsink` package, please first read the [CONTRIBUTING](.github/CONTRIBUTING.md).  In short, contributions are happily 
accepted either via suggestions in the 
[Issues](https://github.com/USEPA/nsink/issues) or via pull request.
