---
title: "Introduction to `nsink`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{intro}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: nsink.bib
csl: ecology.csl
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  cache = TRUE,
  tidy = FALSE
)
load("intro_vignette.rda")
```

```{r setup, echo=FALSE}
library(nsink)
```

The package `nsink` implements an approach to estimate relative nitrogen removal
along a flow path.  This approach is detailed in Kellogg et al. 
[-@kellogg2010geospatial] and builds on peer-reviewed literature in the form of 
reviews and meta-analyses [i.e., @mayer2007meta; @alexander2007role; 
@seitzinger2006denitrification] to estimate nitrogen (N) removal within three 
types of landscape sinks -- wetlands, streams and lakes -- along any given flow 
path within a HUC12 basin. The `nsink` package implements this approach, using 
publicly available spatial data to identify flow paths and estimate N removal in
landscape sinks. Removal rates depend on retention time, which is influenced by 
physical characteristics identified using publicly available spatial data -- 
National Hydrography Dataset (NHD), Watershed Boundary Dataset (WBD), National 
Land Cover Dataset (NLCD), and Soil Survey Geographic Dataset (SSURGO). Static 
maps of a specified HUC-12 basin are generated -- N Removal Efficiency, N 
Loading Index, N Transport Index, and N Delivery Index. These maps may be used 
to inform local decision-making by highlighting areas that are more prone to N 
"leakiness" and areas that contribute to N removal.

The `nsink` packages provides several functions to set up and run an N-Sink 
analysis for a specified 12-digit HUC code.  All required data are downloaded, 
prepared for the analysis, HUC-wide nitrogen removal calculated, and flow paths 
summarized.  Additionally, a convenience function that will run all of the 
required functions for a specified HUC is included.  Details on each step are 
outlined in this vignette.

# Get data
The first step in the N-Sink process is to acquire data needed for the analysis.
The `nsink` package utilizes openly available data from several U.S. Federal 
Government sources.  Each datasets use a 12-digit HUC ID number to select the 
data for download.  The first step is to identify the HUC ID and then download 
the data.

To identify the HUC ID you can use the `nsink_get_huc_id()` function which will 
use a 12-digit HUC name to search all HUCs.  Matches are returned as a data 
frame and an option to return partial or exact matches.

```{r get_huc}
# Get HUC ID - Palmer showing multiple matches
nsink_get_huc_id("Palmer")

# The Niantic
niantic_huc_id <- nsink_get_huc_id("Niantic River")$huc_12
niantic_huc_id
```

With the HUC ID in hand we can now use the `nsink_get_data()` function to 
download the required data.  All data are from publicly available sources and
as of `r lubridate::today()` no authentication is required to access these 
sources.  The HUC ID is required and users may specify a path for storing the 
data as well as indicate whether or not to download the data again if they 
already exist in the data directory.


```{r get_data_norun, eval=FALSE}
# Get data for selected HUC
niantic_download <- nsink_get_data(niantic_huc_id, 
                                   ata_dir = "nsink_niantic_data")
```

In addition to download the data, the function returns the basic information about your download: HUC ID and download location.


```{r get_data_return}
niantic_download
```

# Prepare the data
Once the data is downloaded there are several additional data processing steps 
that are required to subset the data just to the HUC and set all data to a 
common coordinate reference system (CRS).  

These include:

- filter out the HUC boundary
- mask all other data sets to the HUC boundary
- convert all columns names to lower case
- create new columns
- harmonize raster extents
- set all data to common CRS

The `nsink_prep_data()` function will complete all of these processing steps.  
It requires a HUC ID, a specified CRS, and a path to a data directory.  It 
returns a list with all required data for subsequent N-Sink analyses.

A quick note on the CRS.  In the near futre, the preferred way to specify the CRS values will either be with Well-Known Text (WKT) or [EPSG Codes](http://spatialreference.org/ref/epsg/). Proj.4 strings will eventually be deprecated.  Currently the packages that `nsink` relies on are at different stages in implementing the changes to PROJ.  Currently Proj.4 strings are preferred, but that will change soon.

```{r prep_data, eval=FALSE}
# PROJ.4 string for Albers Equal Area
aea <- "+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=23 +lon_0=-96 +x_0=0 +y_0=0 
+ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs"

# Prep data for selected HUC
niantic_data <- nsink_prep_data(niantic_huc_id, projection = aea, 
                data_dir = "nsink_niantic_data")
```

# Calculate removal

The next step in the N-Sink process is to calculate relative nitrogen removal.  
Details on how the nitrogen removal estimates are calculated are available in 
Kellogg et al. [-@kellogg2010geospatial]. The `nsink_calc_removal()` function 
takes the prepared data as an input and returns a list with three items:

- raster_method: This item contains a raster based approach to calculating 
removal.  Used for the static maps of removal.
- land_removal: This represents land based nitrogen removal which is hydric 
soils with areas of impervious surface removed.
- network_removal: This contains removal along the NHD Plus flow network.  
Removal is calculated separately for streams and waterbodies (e.g. lakes and 
reservoirs).

```{r calc_removal, eval=FALSE}
# Calculate removal from prepped data
niantic_removal <- nsink_calc_removal(niantic_data)
```

# Generate and summarize flowpaths

A useful part of the N-Sink approach is not just the calculation of relative 
removal for individual components of the landscape, it is the ability to 
summarize that removal along the legnth of a specified flowpath.  Th `nsink` 
package provides two functions that facilitate this process.  The 
`nsink_generate_flowpath()` function takes a point location as an `sf` object 
and the prepped data (generated by `nsink_prep_data()`) as input and returns an 
`sf` LINESTRING of the flowpath starting from the input point location and 
terminating at the furthest downstream location in the input NHD Plus.  The 
flowpath on land is generated from a flow direction grid.  Once that flowpath 
intersects the stream network, flow is determined by flow along the NHD Plus 
stream network.  First, create the `sf` POINT object.         

```{r flowpath_start}
# Load up the sf package
library(sf)
# Starting point
pt <- c(1948121, 2295822)
start_loc <- st_sf(st_sfc(st_point(c(pt)), crs = aea))
```

You may also determine your point location interactively by plotting your data 
and using the `locator()` function .  First create a simple plot.

```{r flowpath_plot, eval=FALSE}
# Create a simple plot
plot(st_geometry(niantic_data$huc))
plot(st_geometry(niantic_data$lakes), add = T, col = "darkblue")
plot(st_geometry(niantic_data$streams), add = T, col = "blue")
```

With the map made, you can use that to interactively select a location and use 
the x and y to create the `sf` POINT object.

```{r interactive_loc, eval=FALSE}
# Select location on map for starting point
pt <- unlist(locator(n = 1))
# convert to sf POINT
start_loc_inter <- st_sf(st_sfc(st_point(pt), crs = aea))
```

With a point identified, we can use that as the starting location for our 
flowpath.

```{r generate_flowpath, eval=FALSE}
niantic_fp <- nsink_generate_flowpath(start_loc, niantic_data)
```

The reutrned value has both the `flowpath_ends`, the portion of the flowpath on
the land which is created using the flow direction grid, and the 
`flowpath_network` which is the portion of the flowpath from the NHD Plus 
network that occur after the upstream `flowpath_ends` intersect the network.

```{r thefp}
niantic_fp
```

With a flowpath generated we can summarize the relative nitrogen removal along 
that flowpath with the `nsink_summarize_flowpath()` function. It takes the 
flowpath and removal as input.  A data frame is returned with each segment 
identified by type, the percent removal associated with that segment, and 
relative removal.  Total relative removal is 100 - minimum of the `n_out` 
column.

```{r summarize_it}
niantic_fp_removal <- nsink_summarize_flowpath(niantic_fp, niantic_removal)
niantic_fp_removal
100-min(niantic_fp_removal$n_out)
```

# Static maps

Individual flow paths are useful for specific applications, but often it is more
useful to look at removal patterns across the landscape.  The 
`nsink_generate_static_maps()` function provides these HUC wide rasters.  
Required inputs are the prepped data, removal raster, and sampling density.   
The function returns four separate rasters.

- `removal_effic`: Landscape wide estimate of relative nitrogen removal 
percentage.
- `loading_idx`: An index of relative nitrogen loads by land cover class derived 
from published sources
- `transport_idx`: This is relative nitrogen transport for a sample of all 
possible flowpaths in a given HUC.  This is an expensive computational task, so 
`nsink` generates a removal hotspot map based on a sample of flowpaths and the 
final hotspot map is interpolated from these samples and referred to as the 
nitrogen transport index.  The `samp_density` argument controls the number of 
sample flowpaths generated.  
- `delivery_idx`: The delivery index is the combination of the loading index and 
the transport index  It represents which areas of the landscape are 
delivering the most nitrogen to the outflow of the watershed.

```{r static_maps, eval=FALSE}
niantic_static_maps <- nsink_generate_static_maps(niantic_data, niantic_removal, 
                                                  900)
```

And with these static maps made, you can plot them quickly with 
`nsink_plot_removal()`, `nsink_plot_delivery()`, or `nsink_plot_transport()`.  
An example of `nsink_plot_transport()` is below.

```{r plots}
nsink_plot_transport(niantic_static_maps$transport_idx)
```

# Convenience function: Build it all!

The workflow described above includes all the basic functionality.  Some users 
may wish to use `nsink` to calculate the base layers for an N-Sink analysis and 
then build an application outside of R.  A convenience function that downloads 
all data, prepares, calculates removal, and generates static maps has been 
included to facilitate this type of analysis.  The `nsink_build()` function 
requires a HUC ID, coordinate reference system, and sampling density.  An output
folder is also needed but has a default location.  Optional arguments for 
forcing a new download and playing a sound to signal when the build has 
finished. Nothing returns to R, but all prepped data files and `.tif` files are 
written to into the output folder for use in other applications.


```{r build, eval=FALSE}
niantic_huc <- nsink_get_huc_id("Niantic River")$huc_12
aea <- "+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=23 +lon_0=-96 +x_0=0 +y_0=0 
+ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs"
nsink_build(niantic_huc, aea, samp_dens = 900)
```

# References
